Are translation equivalents special? Evidence from simulations and empirical data from bilingual infants

### 2021-12-22 paper/paper_TE_bilingual_vocabulary_model.Rmd
updated paper

The model takes four main parameters: the number of words produced in the dominant language (DOM), the number of words produced in the non-dominant language (NONDOM), vocabulary size of potentially learnable words in each language (LEARNABLE), and a bias parameter (BIAS) which indicates whether the model is biased towards (BIAS > 1) or against (BIAS < 1) learning translation equivalents. The language in which a child knows more words is the dominant language, whereas the one in which a child knows fewer words is the non-dominant language. Next, we turn to the LEARNABLE parameter (i.e., the number of potentially learnable words). If DOM and NONDOM are measured with an instrument such as the MacArthur-Bates Communicative Development inventories (CDI; Fenson et al., 2007), one option would be to set LEARNABLE to be the total number of items on the CDI. For convenience, consider the effect of setting LEARNABLE to 600, as a round number (the actual number of CDI items is usually slightly higher than 600, depending on the language of the adaptation). However, not all children will be capable of learning all of the CDI words. Typically, words that are more concrete, occur more frequently in the input, and words that are associated with babies tend to be acquired earlier than other words (Braginsky et al., 2019), and very young children would not be expected to produce many of the “harder” words on the CDI, such as “lawn mower,” “sidewalk,” or “vitamins”. Thus, a reasonable solution might be to determine how many CDI words are potentially learnable given the child’s developmental level, which could be approximated by their age. Since there are no CDI norms for French-English bilingual children, we will take the average English and French CDI norms to approximate how many words children might lexicalize in either language. For example, imagine that Jamie who is 18 months old produces 50 English words and 20 French words, thus a total of 70 words. Monolingual children his age with the very largest productive vocabularies (those at the 90th percentile averaging between English and French norms) produce a total of 240 words (retrieved from the Wordbank database version 0.3.1; Frank et al., 2016). Although there is likely considerable individual variability as to the cognitive capacity even amongst children of the same age, we argue that averaging between the English and French CDIs at the 90th percentile provides a reasonable — if imperfect — estimate of the number of learnable words (LEARNABLE) that a bilingual child of Jamie’s age could potentially acquire in each language. Thus, we might expect that Jamie could potentially have learned up to 240 words in English and 240 words in French, although he has thus far only learned 50 in English and 20 in French. Note that we model learnability in terms of the number of words expected to be learnable at a particular age, rather than modeling the learnability of individual words at a particular age — a simplifying choice that we will return to in the discussion section.

Using the mathematical concept of independence, we can then quantify the number of translation equivalents (TE) expected given children’s vocabulary sizes in the dominant (DOM) and non-dominant (NONDOM) languages, as well as the number of potentially learnable words (LEARNABLE). If dominant-language and non-dominant-language words are learned independently from each other, we multiply DOM × NONDOM (the number of words known in the dominant and non-dominant language respectively), and divide by the total population of learnable words in one language (LEARNABLE) — which is the possible number of words that could overlap across both languages — to predict the number of translation equivalents. We further introduce the bias parameter (BIAS), which allows us to examine whether translation equivalent learning is best described by the Avoidance, Preference, or Neutral Account. Adding this parameter, translation equivalents can be derived from TE = BIAS × (DOM × NONDOM)/LEARNABLE. For the Avoidance Account, BIAS will be less than 1, meaning that translation equivalents are less easily learned than singlets; for the Preference Account, BIAS will be greater than 1, meaning that translation equivalents are more easily learned than singlets; for the Neutral Account, BIAS is exactly 1 (i.e., the model is unbiased with respect to whether translation equivalents are more difficult or easier to acquire than singlets). Going back to the example of 18-month-old Jamie, we would set the denominator at 240 which is the number of potentially learnable words at 18 months. If translation equivalents are half as easy to learn as singlets (following the Avoidance Account), we would expect Jamie to produce .5×(50×20/240) = 2.1 translation equivalents. Conversely, if translation equivalents are twice as easy to learn as singlets (following the Preference Account), we would expect Jamie to produce 2×(50×20/240) = 8.3 translation equivalents. Under the Neutral Account, we would expect Jamie to learn 1×(50×20/240) = 4.2 translation equivalents.

@@

We present three simulations generated by creating data points for hypothetical children by varying different parameters of the Bilingual Vocabulary Model, in order to explore expected patterns of translation equivalent learning. The number of children in each simulation was fully determined by the number of levels varied for each parameter: each combination of parameters was included, except for those that violated constraints in the model (e.g., NONDOM could not be larger than DOM). In the first simulation, we examined how translation equivalent learning relates to vocabulary balance (BALANCE), as well as different metrics of vocabulary size, including dominant-language vocabulary (DOM), non-dominant language vocabulary (NONDOM), and word vocabulary (WORD). In the second simulation, we explored relations between translation equivalents (TE), balance (BALANCE), and learnable words (LEARNABLE). In the first two simulations, the BIAS parameter was held constant at 1 (Neutral Account); in the third simulation, we varied the bias parameter (BIAS) to compare translation equivalent learning under the Avoidance, Preference, and Neutral Accounts. A summary of the parameter values used in each simulation is provided in Table 2.

@@

landscape(
  kable(table2,
        col.names = c("Simulation",
                      "Learnable words (LEARNABLE)",
                      "Words in dominant Language (DOM)",
                      "Words in non-dominant language (NONDOM)",
                      "Word vocabulary (WORD)",
                      "Balance of vocabulary (BALANCE)",
                      "Bias parameter (BIAS)",
                      "Number of simulated children"),
        align = "clllllll",
        format="latex", booktabs = TRUE,
        caption = "Table 2. Summary of the parameters used in each simulation.") %>%
    column_spec(1:2, width = "45px") %>%
    column_spec(3:7, width = "75px") %>%
    column_spec(8, width = "45px") %>%
    kable_styling(font_size = 8)
)

@@

## 1.1 Simulation 1: Children of the same developmental level with different word vocabularies and balances of vocabulary
In Simulation 1, we first illustrate the relationships between different variables in the model by simulating three hypothetical children who are at the same developmental level and thus have the same number of potentially learnable words (LEARNABLE), but with different word vocabularies (WORD) and BALANCE. For convenience, we set LEARNABLE = 600 in this example, which roughly corresponds to what is expected for an English-French 27-month-old bilingual child (i.e., the most verbal 27-month-old English-learner at the 90th percentile of vocabulary produces around 633 words and the most verbal 27-month-old French-learner produces around 580 words as retrieved from the Wordbank database, giving us an average of 606 words across the two languages; Frank et al., 2016). We set BIAS to 1, meaning that in these examples translation equivalents are equally easy to learn as singlets.

We first illustrate with three simulated children. Infant Annie (small vocabulary, unbalanced exposure) produces 270 words in the dominant language and 30 words in her non-dominant language. She has a word vocabulary of 300, and a balance score of .10 (10% of her words are in the non-dominant language). Based on the formula TE = DOM×NONDOM/LEARNABLE (we drop BIAS from the formula since it is 1 here) and as seen in Table 3, Annie is expected to produce 13.5 translation equivalents. Infant Bernie (small vocabulary, balanced exposure) produces 180 dominant-language words, and 120 non-dominant language words. Like Annie, he has a word vocabulary of 300, but he has a higher balance score of .40 (40% of his words are in the non-dominant language). Based on our formula, we expect Bernie to produce 36 translation equivalents. Comparing Annie and Bernie, two children who produce the same word vocabulary (i.e., WORD is held constant), the child with more balanced language vocabulary (Bernie) is expected to produce more translation equivalents. Like Bernie, infant Charlie also has a balanced vocabulary, but has a larger word vocabulary (WORD), producing 540 words in the dominant language (DOM) and 360 in the non-dominant language (NONDOM) for a total of 900 words (WORD), and thus BALANCE = .40. Based on our formula for Simulation 1, we expect Charlie to produce 324 translation equivalents (TE). Infants Bernie and Charlie illustrate that for two children equal in BALANCE, the child with larger word vocabulary (WORD) is expected to produce more translation equivalents (TE). Other vocabulary metrics are calculated for each hypothetical child as described in Table 3.

We then broadened this simulation to the more general case and examined patterns of translation equivalent learning, where simulated children had the capacity to learn 600 words (LEARNABLE held constant at 600), and their vocabulary size in each language (DOM and NONDOM) varied. BIAS was once again constant at 1. Data from a total of 216 simulated children were generated (see Table 2 for a summary of the parameter values used in this simulation). Based on these values, we derived simulated children’s word vocabulary (WORD, calculated as DOM+NONDOM) and their vocabulary balance (BALANCE, calculated as NONDOM/(DOM+NONDOM)). In Figure 1, we plotted TE knowledge as a function of DOM, NONDOM, and WORD at different levels of BALANCE. Across all three Panels (1A, 1B, and 1C), simulated children with the most balanced vocabulary consistently produced more translation equivalents than other children. Moreover, Panels 1A and 1C show that, as the number of DOM (dominant language words) and WORD (word vocabulary) increased, TE also increased regardless of BALANCE. Interestingly, Panel 1B shows that NONDOM and TE were extremely tightly coupled. In sum, we observed three important patterns, which served as Prediction Set 1 from the Bilingual Vocabulary Model for Study 2:

@@

landscape(
  kable(table3,
      align = "llllll",
      format="latex", booktabs = TRUE,
      caption = "Table 3. Examples for Simulation 1 of three simulated children with different hypothetical word vocabularies (WORD) and vocabulary balance (BALANCE), where the number of learnable words (LEARNABLE) = 600 and BIAS = 1. Vocabulary values are rounded to the nearest integer.") %>%
  pack_rows("Main Parameters", 1, 4) %>%
  pack_rows("Derived Parameters", 5, 10) %>%
  column_spec(c(1), width = "100px") %>%
  column_spec(c(2:3), width = "140px") %>%
  column_spec(c(4:6), width = "60px") %>%
  kable_styling(font_size = 8) 
)
```

@@

## visualizing Simulation 1
simulation1 <- data_simulation1 %>%
  # convert to long data
  pivot_longer(c(WORD, DOM, NONDOM), names_to = "vocab_type", values_to = "number") %>%
  mutate(vocab_type = factor(vocab_type, levels=c("WORD", "DOM", "NONDOM"))) %>%
  mutate(vocab_type = recode(vocab_type, 
                             WORD = "Panel 1A: \n Total vocabulary \n (WORD)",
                             DOM = "Panel 1B: \n Dominant vocabulary \n (DOM)", 
                             NONDOM = "Panel 1C: \n Non-dominant vocabulary \n (NONDOM)")) %>%
  ggplot(aes(x = number, y = TE, color = bin_BALANCE, linetype = bin_BALANCE)) +
  stat_smooth(method = lm, se = F) +
  scale_color_manual(values=c("#bdbdbd", "#737373", "#525252", "#252525", "#000000")) + 
  scale_linetype_manual(values=c("longdash", "dotdash", "dashed", "twodash", "solid")) +
  theme_minimal() + 
  facet_grid(. ~ vocab_type) +
  theme(axis.text.x = element_text(angle = 90),
        legend.position = "bottom",
        plot.title = element_text(face = 'bold', size = 12, hjust = 0, vjust = 2),
        plot.title.position = "plot",
        text = element_text(size=12),
        panel.spacing.x = unit(2, "lines"),
        panel.background = element_rect(fill = NA, color = "black")) +
  ylim(0, 600) +
  xlim(0, 1000) +
  labs(x = "Number of words", 
       y = "Number of translation equivalents (TE)", 
       color = "Vocabulary balance (BALANCE)",
       linetype = "Vocabulary balance (BALANCE)",
       title = "1) Simulated data") +
  guides(color = guide_legend(reverse = T, nrow = 1),
         linetype = guide_legend(reverse = T, nrow = 1)) 
```

```{r simulation1-plus_observed_data, echo = FALSE, cache = TRUE}
Observed_plot_inputs_TE <- keepers_ws_TE %>% 
  select(baby_id, ID_testdate, age_days, number_of_te, word_vocab, total_words_dom, total_words_nondom, balance_vocab) %>%
  pivot_longer(c(word_vocab, total_words_dom, total_words_nondom), names_to = "vocab_type", values_to = "number") %>%
  mutate(vocab_type = factor(vocab_type, levels=c("word_vocab", "total_words_dom", "total_words_nondom"))) %>%
  mutate(vocab_type = recode(vocab_type, 
                             word_vocab = "Panel 2A: \n Total vocabulary \n (WORD)",
                             total_words_dom = "Panel 2B: \n Dominant vocabulary \n (DOM)", 
                             total_words_nondom = "Panel 2C:\n Non-dominant vocabulary \n (NONDOM)")) %>%
  mutate(bin_BALANCE = cut(balance_vocab, breaks = c(0, 0.1, 0.2, 0.3, 0.4, 0.5), include.lowest=T,
                     labels = c("0.1", "0.2", "0.3", "0.4", "0.5"))) %>%
  ggplot(aes(x = number, y = number_of_te, color = bin_BALANCE, linetype = bin_BALANCE)) +
  #geom_point() +
  stat_smooth(method = lm, se = F) +
  scale_color_manual(values=c("#bdbdbd", "#737373", "#525252", "#252525", "#000000")) + 
  scale_linetype_manual(values=c("longdash", "dotdash", "dashed", "twodash", "solid")) + 
  theme_minimal() + 
  facet_grid(. ~ vocab_type) +
  theme(axis.text.x = element_text(angle = 90),
        plot.title = element_text(face = 'bold', size = 12, hjust = 0, vjust = 2),
        plot.title.position = "plot",
        text = element_text(size=12),
        legend.position = "bottom",
        panel.spacing.x = unit(2, "lines"),
        panel.background = element_rect(fill = NA, color = "black")) +
  ylim(0, 600) +
  xlim(0, 1000) +
  labs(x = "Number of words", 
       y = "Number of translation equivalents (TE)", 
       color = "Vocabulary balance (BALANCE)",
       linetype = "Vocabulary balance (BALANCE)",
       title = "2) Observed data") +
  guides(colour = guide_legend(reverse = T, nrow = 1),
         linetype = guide_legend(reverse = T, nrow = 1))

@@

Translation equivalent knowledge was simulated across children at six developmental levels (the number of LEARNABLE words = 100, 200, 300, 400, 500, 600), in conjunction with a wide range of values for words in the dominant language (DOM) and the non-dominant language (NONDOM). In total, data from 301 simulated children were generated (see Table 2 for a summary of the parameters used in this simulation). Again, balance (BALANCE) was calculated based on the values of DOM and NONDOM. We also calculated the number of singlet words in the dominant (DOM-SINGLET) and non-dominant (NONDOM-SINGLET) languages by the formula DOM-SINGLET = DOM - TE and NONDOM-SINGLET = NONDOM - TE. Figure 2 plots the number of CONCEPT, TE, DOM-SINGLET, and NONDOM-SINGLET at different BALANCE levels across different levels of LEARNABLE. As shown in Panel 1A, simulated children at a later developmental level had larger concept vocabularies (CONCEPT). In Panel 1B, we continued to observe a pattern reported in prediction 1a, whereby simulated children with more balanced vocabularies produced more translation equivalents (TE). Moreover, regardless of balance, simulated children at later developmental levels (i.e., older children with more potentially LEARNABLE words) acquired more translation equivalents (TE). Panels 1C and 1D show that simulated children at later developmental levels also produced more singlets in the dominant and non-dominant languages, and that more balanced children would produce more singlets in their non-dominant language, but fewer singlets in their dominant language than less balanced children. Overall, we generated 4 additional predictions (Prediction Set 2) made by the Bilingual Vocabulary Model. Compared to children at an earlier developmental level (i.e., younger infants with fewer potentially learnable words), children at a later developmental level (i.e., older infants with more potentially learnable words) will

@@

## visualizing Simulation 2
simulation2 <- data_simulation2 %>%
  pivot_longer(-c(LEARNABLE, WORD, BALANCE, bin_BALANCE), names_to = "vocab_type", values_to = "number") %>%
  filter(vocab_type %notin% c("DOM", "NONDOM")) %>%
  mutate(vocab_type = factor(vocab_type, levels=c("CONCEPT", "TE", "DOM_SINGLET", "NONDOM_SINGLET")),
         vocab_type = recode(vocab_type, 
                             CONCEPT = "Panel 1A: \n Concept vocabulary \n (CONCEPT)",
                             TE = "Panel 1B: \n Translation equivalents \n (TE)",
                             DOM_SINGLET = "Panel 1C: Singlets in \n dominant language \n (DOM_SINGLET)",
                             NONDOM_SINGLET = "Panel 1D: Singlets in \n non-dominant language \n (NONDOM_SINGLET)")) %>%
  ggplot(aes(x = LEARNABLE, y = number, color = bin_BALANCE, linetype = bin_BALANCE)) +
  facet_grid(. ~ vocab_type) +
  stat_smooth(method = lm, se = F) + 
  scale_color_manual(values=c("#bdbdbd", "#737373", "#525252", "#252525", "#000000")) + 
  scale_linetype_manual(values=c("longdash", "dotdash", "dashed", "twodash", "solid")) +
  theme_minimal() +
  theme(legend.position = "bottom",
        plot.title = element_text(face = 'bold', size = 10, hjust = 0, vjust = 2),
        plot.title.position = "plot",
        text = element_text(size=10),
        panel.spacing.x = unit(2, "lines"),
        panel.background = element_rect(fill = NA, color = "black")) + 
  guides(fill=guide_legend(nrow=3,byrow=TRUE)) + 
  labs(x = "\nNumber of learnable words (LEARNABLE)", 
       y = "Number of concepts (CONCEPT)", 
       color = "Vocabulary balance (BALANCE)",
       linetype = "Vocabulary balance (BALANCE)",
       title = "1) Simulated data") +
  guides(color = guide_legend(reverse = T, nrow = 1),
         linetype = guide_legend(reverse = T, nrow = 1)) 
```

>>>>

The model takes four main parameters: the number of words produced in the dominant language (DOM), the number of words produced in the non-dominant language (NONDOM), vocabulary size of potentially learnable words in each language (LEARNABLE), and a bias parameter (BIAS) which indicates whether the model is biased towards (BIAS > 1) or against (BIAS < 1) learning translation equivalents. The language in which a child knows more words is the dominant language, whereas the one in which a child knows fewer words is the non-dominant language. Next, we turn to the LEARNABLE parameter (i.e., the number of potentially learnable words). If DOM and NONDOM are measured with an instrument such as the MacArthur-Bates Communicative Development inventories (CDI; Fenson et al., 2007), one option would be to set LEARNABLE to be the total number of items on the CDI. For convenience, consider the effect of setting LEARNABLE to 600, as a round number (the actual number of CDI items is usually slightly higher than 600, depending on the language of the adaptation). However, not all children will be capable of learning all of the CDI words. Typically, words that are more concrete occur more frequently in the input, and words that are associated with babies tend to be acquired earlier than other words (Braginsky et al., 2019). That is, very young children would not be expected to produce many of the advanced words on the CDI, such as “lawn mower,” “sidewalk,” or “vitamins”. Thus, a reasonable solution might be to determine how many CDI words are potentially learnable given the child’s developmental level, which could be approximated by their age. Since there are no CDI norms for French—English bilingual children, we will take the average English and French CDI norms to approximate how many words children might lexicalize in either language. For example, imagine that Jamie who is 18 months old produces 50 English words and 20 French words, thus a total of 70 words. Monolingual children his age with the very largest productive vocabularies (those at the 90th percentile averaging between English and French norms) produce a total of 240 words (retrieved from the Wordbank database version 0.3.1; Frank et al., 2016). Although there is likely considerable individual variability as to the cognitive capacity even amongst children of the same age, we argue that averaging between the English and French CDIs at the 90th percentile provides a reasonable — if imperfect — estimate of the number of learnable words (LEARNABLE) that a bilingual child of Jamie’s age could potentially acquire in each language. Thus, we might expect that Jamie could potentially have learned up to 240 words in English and 240 words in French, although he has thus far only learned 50 in English and 20 in French. Note that we model learnability in terms of the number of words expected to be learnable at a particular age, rather than modeling the learnability of individual words at a particular age — a simplifying choice that we will return to in the discussion section.

Using the mathematical concept of independence, we can then quantify the number of translation equivalents (TE) expected given children’s vocabulary sizes in the dominant (DOM) and non-dominant (NONDOM) languages, as well as the number of potentially learnable words (LEARNABLE). If dominant-language and non-dominant-language words are learned independently from each other, we multiply DOM × NONDOM (the number of words known in the dominant and non-dominant language respectively), and divide by the total population of learnable words in one language (LEARNABLE) — which is the possible number of words that could overlap across both languages — to predict the number of translation equivalents. We further introduce the bias parameter (BIAS), which allows us to examine whether translation equivalent learning is best described by the Avoidance, Preference, or Neutral Account. Adding this parameter, translation equivalents can be derived from TE = BIAS × (DOM × NONDOM)/LEARNABLE. For the Avoidance Account, BIAS will be less than 1, meaning that translation equivalents are less easily learned than singlets; for the Preference Account, BIAS will be greater than 1, meaning that translation equivalents are more easily learned than singlets; for the Neutral Account, BIAS is exactly 1 (i.e., the model is unbiased with respect to whether translation equivalents are more difficult or easier to acquire than singlets). Going back to the example of 18-month-old Jamie, we would set the denominator at 240 which is the number of potentially learnable words at 18 months. If translation equivalents are half as easy to learn as singlets (following the Avoidance Account), we would expect Jamie to produce .5×(50×20/240) = 2.1 translation equivalents. Conversely, if translation equivalents are twice as easy to learn as singlets (following the Preference Account), we would expect Jamie to produce 2×(50×20/240) = 8.3 translation equivalents. Under the Neutral Account, we would expect Jamie to produce 1×(50×20/240) = 4.2 translation equivalents.

@@

We present three simulations generated by creating data points for hypothetical children by varying different parameters of the Bilingual Vocabulary Model, in order to explore expected patterns of translation equivalent learning. The number of children in each simulation was fully determined by the number of levels varied for each parameter: each combination of parameters was included, except those that violated constraints in the model (e.g., NONDOM could not be larger than DOM). In the first simulation, we examined how translation equivalent learning relates to vocabulary balance (BALANCE), as well as different metrics of vocabulary size, including dominant-language vocabulary (DOM), non-dominant language vocabulary (NONDOM), and word vocabulary (WORD). In the second simulation, we explored relations between translation equivalents (TE), balance (BALANCE), and learnable words (LEARNABLE). In the first two simulations, the BIAS parameter was held constant at 1 (Neutral Account); in the third simulation, we varied the BIAS parameter to compare translation equivalent learning under the Avoidance, Preference, and Neutral Accounts. A summary of the parameter values used in each simulation is provided in Table 2.

@@

landscape(
  kable(table2,
        col.names = c("Simulation",
                      "Learnable words (LEARNABLE)",
                      "Words in dominant Language (DOM)",
                      "Words in non-dominant language (NONDOM)",
                      "Word vocabulary (WORD)",
                      "Balance of vocabulary (BALANCE)",
                      "Bias parameter (BIAS)",
                      "Number of simulated children"),
        align = "clllllll",
        format="latex", booktabs = TRUE,
        caption = "Table 2. Summary of the parameters used in each simulation.") %>%
    column_spec(1:2, width = "45px") %>%
    column_spec(3:7, width = "75px") %>%
    column_spec(8, width = "45px") %>%
    kable_styling(font_size = 8) %>%
    footnote(general = "The number of children simulated is derived from the total number of allowed combinations of the parameters (i.e., excluding combinations that violate parameter constraints; for example NONDOM could not be larger than DOM).",
             threeparttable = T)
)

@@

## 1.1 Simulation 1: Children of the same developmental level with different word vocabularies and balances of vocabulary
In Simulation 1, we first illustrate the relationships between different variables in the model by simulating three hypothetical children who are at the same developmental level and thus have the same number of potentially learnable words (LEARNABLE), but with different word vocabularies (WORD) and BALANCE. For convenience, we set LEARNABLE = 600 in this example, which roughly corresponds to what is expected for an English–French 27-month-old bilingual child (i.e., the most verbal 27-month-old English-learner at the 90th percentile of vocabulary produces around 633 words and the most verbal 27-month-old French-learner produces around 580 words as retrieved from the Wordbank database, giving us an average of 606 words across the two languages; Frank et al., 2016). We set BIAS to 1, meaning that in these examples translation equivalents and singlets are equally easy to learn.

Our first simulated infant, Annie (small vocabulary, unbalanced exposure), produces 270 words in the dominant language and 30 words in her non-dominant language. She has a word vocabulary of 300, and a balance score of .10 (10% of her words are in the non-dominant language). Based on the formula TE = DOM×NONDOM/LEARNABLE (we drop BIAS from the formula since it is 1 here) and as seen in Table 3, Annie is expected to produce 13.5 translation equivalents. Our second infant, Bernie (small vocabulary, balanced exposure), produces 180 dominant-language words and 120 non-dominant language words. Like Annie, he has a word vocabulary of 300, but he has a higher balance score of .40 (40% of his words are in the non-dominant language). Based on our formula, we expect Bernie to produce 36 translation equivalents. Comparing Annie and Bernie, two children who produce the same total number of word vocabulary (i.e., WORD is held constant), the child with the more balanced bilingual vocabulary (Bernie) is expected to produce more translation equivalents. Like Bernie, our third infant Charlie also has a balanced vocabulary, but has a larger word vocabulary (WORD), producing 540 words in the dominant language (DOM) and 360 in the non-dominant language (NONDOM), for a total of 900 words (WORD), and thus BALANCE = .40. Based on our formula for Simulation 1, we expect Charlie to produce 324 translation equivalents (TE). Infants Bernie and Charlie illustrate that for two children equal in BALANCE, the child with larger word vocabulary (WORD) is expected to produce more translation equivalents (TE). Other vocabulary metrics are calculated for each hypothetical child as described in Table 3.

We then broadened this simulation to the more general case and examined patterns of translation equivalent learning, where simulated children had the capacity to learn 600 words (LEARNABLE held constant at 600), and their vocabulary size in each language (DOM and NONDOM) varied. BIAS remained constant at 1. Data from a total of 216 simulated children were generated (see Table 2 for a summary of the parameter values used in this simulation). Based on these values, we derived simulated children’s word vocabulary (WORD, calculated as DOM+NONDOM) and their vocabulary balance (BALANCE, calculated as NONDOM/(DOM+NONDOM)). In Figure 1, we plotted TE knowledge as a function of DOM, NONDOM, and WORD at different levels of BALANCE. Across all three Panels (1A, 1B, and 1C), simulated children with the most balanced vocabulary consistently produced more translation equivalents than other children. Moreover, Panels 1A and 1C show that, as DOM (dominant language words) and WORD (word vocabulary) increased, TE (translation equivalents) also increased regardless of BALANCE. Interestingly, Panel 1B shows that NONDOM and TE were extremely tightly coupled. In sum, we observed three important patterns, which served as Prediction Set 1 from the Bilingual Vocabulary Model for Study 2:

@@

landscape(
  kable(table3,
      align = "llllll",
      format="latex", booktabs = TRUE,
      caption = "Table 3. Examples for Simulation 1 of three simulated children with different hypothetical word vocabularies (WORD) and vocabulary balances (BALANCE), where the number of learnable words (LEARNABLE) = 600 and BIAS = 1. Vocabulary values are rounded to the nearest integer.") %>%
  pack_rows("Main Parameters", 1, 4) %>%
  pack_rows("Derived Parameters", 5, 10) %>%
  column_spec(c(1), width = "100px") %>%
  column_spec(c(2:3), width = "140px") %>%
  column_spec(c(4:6), width = "60px") %>%
  kable_styling(font_size = 8) 
)
```

@@

## visualizing Simulation 1
simulation1 <- data_simulation1 %>%
  # convert to long data
  pivot_longer(c(WORD, DOM, NONDOM), names_to = "vocab_type", values_to = "number") %>%
  mutate(vocab_type = factor(vocab_type, levels=c("WORD", "DOM", "NONDOM"))) %>%
  mutate(vocab_type = recode(vocab_type, 
                             WORD = "Panel 1A: \n Total vocabulary \n (WORD)",
                             DOM = "Panel 1B: \n Dominant vocabulary \n (DOM)", 
                             NONDOM = "Panel 1C: \n Non-dominant vocabulary \n (NONDOM)")) %>%
  ggplot(aes(x = number, y = TE, color = bin_BALANCE, linetype = bin_BALANCE)) +
  stat_smooth(method = lm, se = F) +
  scale_color_manual(values=c("#bdbdbd", "#737373", "#525252", "#252525", "#000000")) + 
  scale_linetype_manual(values=c("longdash", "dotdash", "dashed", "twodash", "solid")) +
  theme_minimal() + 
  facet_grid(. ~ vocab_type) +
  theme(axis.text.x = element_text(angle = 90),
        legend.position = "bottom",
        legend.key.width=unit(1.1,"cm"),
        plot.title = element_text(face = 'bold', size = 12, hjust = 0, vjust = 2),
        plot.title.position = "plot",
        text = element_text(size=12),
        panel.spacing.x = unit(2, "lines"),
        panel.background = element_rect(fill = NA, color = "black")) +
  ylim(0, 600) +
  xlim(0, 1000) +
  labs(x = "Number of words", 
       y = "Number of translation equivalents (TE)", 
       color = "Vocabulary balance (BALANCE)",
       linetype = "Vocabulary balance (BALANCE)",
       title = "1) Simulated data") +
  guides(color = guide_legend(reverse = T, nrow = 1),
         linetype = guide_legend(reverse = T, nrow = 1)) 
```

```{r simulation1-plus_observed_data, echo = FALSE, cache = TRUE}
Observed_plot_inputs_TE <- keepers_ws_TE %>% 
  select(baby_id, ID_testdate, age_days, number_of_te, word_vocab, total_words_dom, total_words_nondom, balance_vocab) %>%
  pivot_longer(c(word_vocab, total_words_dom, total_words_nondom), names_to = "vocab_type", values_to = "number") %>%
  mutate(vocab_type = factor(vocab_type, levels=c("word_vocab", "total_words_dom", "total_words_nondom"))) %>%
  mutate(vocab_type = recode(vocab_type, 
                             word_vocab = "Panel 2A: \n Total vocabulary \n (WORD)",
                             total_words_dom = "Panel 2B: \n Dominant vocabulary \n (DOM)", 
                             total_words_nondom = "Panel 2C:\n Non-dominant vocabulary \n (NONDOM)")) %>%
  mutate(bin_BALANCE = cut(balance_vocab, breaks = c(0, 0.1, 0.2, 0.3, 0.4, 0.5), include.lowest=T,
                     labels = c("0.1", "0.2", "0.3", "0.4", "0.5"))) %>%
  ggplot(aes(x = number, y = number_of_te, color = bin_BALANCE, linetype = bin_BALANCE)) +
  #geom_point() +
  stat_smooth(method = lm, se = F) +
  scale_color_manual(values=c("#bdbdbd", "#737373", "#525252", "#252525", "#000000")) + 
  scale_linetype_manual(values=c("longdash", "dotdash", "dashed", "twodash", "solid")) + 
  theme_minimal() + 
  facet_grid(. ~ vocab_type) +
  theme(axis.text.x = element_text(angle = 90),
        plot.title = element_text(face = 'bold', size = 12, hjust = 0, vjust = 2),
        plot.title.position = "plot",
        text = element_text(size=12),
        legend.position = "bottom",
        legend.key.width=unit(1.1,"cm"),
        panel.spacing.x = unit(2, "lines"),
        panel.background = element_rect(fill = NA, color = "black")) +
  ylim(0, 600) +
  xlim(0, 1000) +
  labs(x = "Number of words", 
       y = "Number of translation equivalents (TE)", 
       color = "Vocabulary balance (BALANCE)",
       linetype = "Vocabulary balance (BALANCE)",
       title = "2) Observed data") +
  guides(colour = guide_legend(reverse = T, nrow = 1),
         linetype = guide_legend(reverse = T, nrow = 1))

@@

Translation equivalent knowledge was simulated across children at six developmental levels (number of LEARNABLE words = 100, 200, 300, 400, 500, 600), in conjunction with a wide range of values for words in the dominant language (DOM) and the non-dominant language (NONDOM). In total, data from 301 simulated children were generated (see Table 2 for a summary of the parameters used in this simulation). Again, balance (BALANCE) was calculated based on the values of DOM and NONDOM. We also calculated the number of singlet words in the dominant (DOM-SINGLET) and non-dominant (NONDOM-SINGLET) languages by the formula DOM-SINGLET = DOM - TE and NONDOM-SINGLET = NONDOM - TE. Figure 2 plots the number of CONCEPT, TE, DOM-SINGLET, and NONDOM-SINGLET at different BALANCE levels across different levels of LEARNABLE. As shown in Panel 1A, simulated children at a later developmental level had larger concept vocabularies (CONCEPT). In Panel 1B, we continued to observe a pattern reported in prediction 1a, whereby simulated children with more balanced vocabularies produced more translation equivalents (TE). Moreover, regardless of balance, simulated children at later developmental levels (i.e., older children with more potentially LEARNABLE words) acquired more translation equivalents (TE). Panels 1C and 1D show that simulated children at later developmental levels also produced more singlets in the dominant and non-dominant languages, and that more balanced children produced more singlets in their non-dominant language, but fewer singlets in their dominant language than less balanced children. Overall, we generated 4 additional predictions (Prediction Set 2) made by the Bilingual Vocabulary Model. Compared to children at an earlier developmental level (i.e., younger infants with fewer potentially learnable words), children at a later developmental level (i.e., older infants with more potentially learnable words) will

@@

## visualizing Simulation 2
simulation2 <- data_simulation2 %>%
  pivot_longer(-c(LEARNABLE, WORD, BALANCE, bin_BALANCE), names_to = "vocab_type", values_to = "number") %>%
  filter(vocab_type %notin% c("DOM", "NONDOM")) %>%
  mutate(vocab_type = factor(vocab_type, levels=c("CONCEPT", "TE", "DOM_SINGLET", "NONDOM_SINGLET")),
         vocab_type = recode(vocab_type, 
                             CONCEPT = "Panel 1A: \n Concept vocabulary \n (CONCEPT)",
                             TE = "Panel 1B: \n Translation equivalents \n (TE)",
                             DOM_SINGLET = "Panel 1C: Singlets in \n dominant language \n (DOM_SINGLET)",
                             NONDOM_SINGLET = "Panel 1D: Singlets in \n non-dominant language \n (NONDOM_SINGLET)")) %>%
  ggplot(aes(x = LEARNABLE, y = number, color = bin_BALANCE, linetype = bin_BALANCE)) +
  facet_grid(. ~ vocab_type) +
  stat_smooth(method = lm, se = F) + 
  scale_color_manual(values=c("#bdbdbd", "#737373", "#525252", "#252525", "#000000")) + 
  scale_linetype_manual(values=c("longdash", "dotdash", "dashed", "twodash", "solid")) +
  theme_minimal() +
  theme(legend.position = "bottom",
        legend.key.width=unit(1.1,"cm"),
        plot.title = element_text(face = 'bold', size = 10, hjust = 0, vjust = 2),
        plot.title.position = "plot",
        text = element_text(size=10),
        panel.spacing.x = unit(2, "lines"),
        panel.background = element_rect(fill = NA, color = "black")) + 
  guides(fill=guide_legend(nrow=3,byrow=TRUE)) + 
  labs(x = "\nNumber of learnable words (LEARNABLE)", 
       y = "Number produced", 
       color = "Vocabulary balance (BALANCE)",
       linetype = "Vocabulary balance (BALANCE)",
       title = "1) Simulated data") +
  guides(color = guide_legend(reverse = T, nrow = 1),
         linetype = guide_legend(reverse = T, nrow = 1)) 
```
